Scheduling and Amortization for MPC

### 2022-02-20 biometric_vectorization_example.tex
Made a plan. Added an overview section.
Reorganized, removed TODOS, added other TODOs.

\subsection{Example: biometric}

@@

This phase removes redundant dimensionality. 
It starts by removing redundant dimensions in MOTION loops followed
by removal of redundant drop dimension statements.
It then does (extended) constant propagation 
to "bypass" raise statements, copy propagation
and dead code elimination.

@@

// j is redundant for sum!3 and sum!4
for j in range(0, D!0):
  sum!3[I] = PHI(sum!2^[I], sum!4[I])       
  sum!4[I] = (sum!3[I] + p!3[I,j])

@@

\ana{TODO: It will be good to add a pass in Phase 3, or post Phase 3, to get rid of extra dimensionality. 
Once we have vectorized as much as possible, there is no need to keep higher 
dimensionality. E.g., there is no need for the 2-dimensional array sum in the first for loop, 
and similarly, there is no need for 1-dimensional arrays for min\_sum and min\_index.
We should get rid of these before generating MOTION code.}

\ana{I don't have a crisp algorithm in mind. I don't think it will be difficult but we have to 
think this through. Put on our TODO list for us to think about since it will simplify MOTION 
code tremendously.}

\ana{Begin Work-in-progress notes on implementation. Just some thoughts...}

For now, I suggest we go with a "naive" implementation and try to do only minimal optimizations. 
All arrays are stored in the program as one-dimensional arrays. Arrays lifted from scalars have dimensionality of the loop enclosure in 
the original code, e.g., \texttt{sum!3} is two dimensional \texttt{[N!0,D!0]} (but represented linearly in row-major order; all these "copies" are 
arrays of references to shares). Naturally, all writes into these arrays are canonical writes, i.e., $A[i,j] = ...$ and no things like $A[i+1,j-1] = ...$.

@@

\begin{verbatim}
// project the (j-1)-st column vector:
A = sum!4[I,j-1]
// project j-th column from p!3: 
B = p!3[I,j] 
// we get a N!0-size vector of references to NEW shares:
C = ADD_SIMD(A,B) 
// Writes j-th column of sum!4 with the new references:
sum!4[I,j] = C 
// may eventually optimize since C is A
\end{verbatim}

There are three kinds of arrays (for now all kept internally as one-dimensional arrays, but that's under discussion).
\begin{itemize}

>>>>

\subsection{Example: Biometric}

@@

This phase removes redundant dimensionality. 
It starts by removing redundant dimensions in MOTION loops followed
by removal of redundant drop dimension statements.
It then does (extended) constant propagation 
to "bypass" raise statements, followed by copy propagation
and dead code elimination.

@@

// j is redundant for sum!3 and sum!4
for j in range(0, D!0):
  sum!3[I] = PHI(sum!2[I], sum!4[I])       
  sum!4[I] = (sum!3[I] + p!3[I,j])

@@

%\ana{TODO: It will be good to add a pass in Phase 3, or post Phase 3, to get rid of extra dimensionality. Once we have vectorized as much as possible, there is no need to keep higher dimensionality. E.g., there is no need for the 2-dimensional array sum in the first for loop, and similarly, there is no need for 1-dimensional arrays for min\_sum and min\_index. We should get rid of these before generating MOTION code.}

%\ana{I don't have a crisp algorithm in mind. I don't think it will be difficult but we have to think this through. Put on our TODO list for us to think about since it will simplify MOTION code tremendously.}

\ana{Begin Work-in-progress notes on implementation. Just some thoughts...} 
\begin{comment}
For now, I suggest we go with a "naive" implementation and try to do only minimal optimizations. 
All arrays are stored in the program as one-dimensional arrays. Arrays lifted from scalars have dimensionality of the loop enclosure in 
the original code, e.g., \texttt{sum!3} is two dimensional \texttt{[N!0,D!0]} (but represented linearly in row-major order; all these "copies" are 
arrays of references to shares). Naturally, all writes into these arrays are canonical writes, i.e., $A[i,j] = ...$ and no things like $A[i+1,j-1] = ...$.

@@

\begin{verbatim}
// project the (j-1)-st column vector:
A = sum!4[I,j-1]
// project j-th column from p!3: 
B = p!3[I,j] 
// we get a N!0-size vector of references to NEW shares:
C = ADD_SIMD(A,B) 
// Writes j-th column of sum!4 with the new references:
sum!4[I,j] = C 
// may eventually optimize since C is A
\end{verbatim}
\end{comment}
There are three kinds of arrays (for now all kept internally as one-dimensional arrays, but that's under discussion).
\begin{itemize}

### 2022-02-20 paper_SIMD.tex
Reorganized, removed TODOS, added other TODOs.

Our compiler produces the following vectorized program. (Note that this is
still higher-level IR, our compiler turns this code into MOTION variables 
loops and SIMD primitives, which MOTION then uses to generate the circuit.)

{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
// S!0^ is same as S!0. C!0 replicates C!0 N-times: 
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

for j in range(0, D!0):
  // sum!2[I], sum!3[I], sum!4[I] are vectors of size N  
  // Computes N intermediate sums "at once"
  sum!3[I] = PHI(sum!2^[I], sum!4[I])       
  sum!4[I] = ADD_SIMD(sum!3[I],p!3[I,j])

@@

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion. \ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item All MPC instructions have the same unit cost, 1 unit. \ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item There is unlimited bandwidth---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely 1 unit.\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions scheduled in parallel benefit from amortization and cost 1, however a MUL and a MUX instructions scheduled
in parallel still cost 2. \ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

@@

@@

@@

We assume arbitrarily nested loops in the MPC-source IR and read-only arrays. We assume that loops range from 0 to some constant N. Arrays are linearized (row-major order as in MOTION) and accesses are via functions of the induction variables of the enclosing loops. We write $i_1,i_2,...i_k$ to denote the loop nest: $i_1$ is the outermost loop, $i_2$, is immediately nested in $i_1$, and so on until $i_k$. As an example, a statement nested in $i_1,i_2,...i_k$ can access array $A[f(i_1,i_2,...i_k)]$. We write $A[i_1,i_2,...i_k]$ interchangeably. \ana{Need to state this more precisely.}

@@

\subsection{Def-use Edges}

@@

\begin{itemize}
\item same-level forward $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest $i_1,i_2,...i_k$. E.g., $\texttt{d} = \mathit{SUM}(\texttt{S}[i,j], \texttt{C}[j])$ to $\texttt{p} = \mathit{MUL}(\texttt{d},\texttt{d})$ in Biometric is a same-level edge. A $\phi$ node can be a source of a same-level forward edge but not a target.
\item outer-to-inner forward $X\rightarrow Y$ where $X$ is in an outer loop nest, $i_1,i_2,...i_j$, and $Y$ is in an inner one, $i_1,i_2,...i_j,...,i_k$. A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer forward $X\rightarrow Y$ where $X$ is a $phi$-node in an inner loop nest, $i_1,i_2,...i_k,i_{k+1}$, and $Y$ is in the enclosing loop nest $i_1,i_2,...i_k$. E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer forward edge. Note that the source is \emph{always} a $\phi$-node in the immediately enclosing loop. The interpretation of this edge is that the use node $Y$ uses the definition made in the last iteration of the inner loop.
\ben{The current representation of pseudo $\phi$-nodes shows them attached to the loop header (i.e. in the loop).  We may want to clarify that they are also evaluated at the loop's termination.}
\item same-level back-edge $X\rightarrow Y$. $Y$ is a phi-node in the header of the loop and $X$ is a definition of the variable in the loop body. E.g., $\texttt{min}_1 = \mathit{MUX}(\texttt{c},\texttt{sum}_1,\texttt{min}_1)$ to $\texttt{min}_0 = \phi(\texttt{min}_1, 10000)$ in Biometric is a same-level back-edge.
\item inner-to-outer back-edge $X\rightarrow Y$: $X$ and $Y$ are both $\phi$-nodes for some variable. The source $X$ is in a loop nested into $Y$'s loop (not necessarily immediately).

@@

\subsection{Helper Functions}

@@

@@

\section{Vectorization}

\ana{Add back-edges into Phase 1. A back-edge from a non-phi-node in loop $i$ to a phi-node in loop $i$'s header is a same-level edge. The only difference with the handling of normal same-level forward def-use is that the operand index will become $i-1$. A back edge from a phi-node in loop $j$ to a phi-node in loop $i$, where $j$ is nested in $i$ is an inner-to-outer edge and will require dropping dimension. We can show these are the only kinds of back-edges that may occur.}

@@

\ishaq{We need to think of a way to handle 4th case for def-use, when a nested block B def is followed by another, different, nested block B' use. Note: The handwritten notes say we can break it into 2 edges. One idea to this end is to insert a def' node which is a copy of the original def. The def' should be placed in a block that contains both B and B'.}
\ishaq{THE ABOVE COMMENT IS RESOLVED. After transformation to SSA, there should be an intermediate variable (phi node) that is assigned a value depending on whether control branched or not. Should we put this into the writeup?}

@@

%\ana{Clarify why this helps. Still work in progress.}
This is to reduce the dimensionality of simd-ified computation. Technically, $\texttt{A}_j = \emph{MUX}(..., \texttt{A}_k, \texttt{A}_l)$ is
a simdified operation that can be carried out in parallel "in one round". However, particularly when \texttt{A} is a multi-dimensional array, there is
substantial increase in the size of the arrays (vectors) we send to SIMD operations. Refining to an update to a specific index would reduce
the size of those vectors. Note that this is a heuristic that handles a common case, but not all cases of array updates.
\ana{TODO: Simplify this algorithm, taking into account the restriction to canonical updates. It should be handling \_all\_ cases.}

@@

\item same-level $X \rightarrow Y$. We do nothing, just propagate the array, which happens to be of the right dimension. \ana{There might be some opportunities to do copy propagation optimization and save some cycles, but let's leave this for later.}

@@

\section{Divide-and-Conquer}

>>>>

Our compiler produces the following vectorized program. (Note that this is
still higher-level IR, our compiler turns this code into MOTION variables, 
loops and SIMD primitives, which MOTION then uses to generate the circuit.)

{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
// S!0^ is same as S!0. C!0^ replicates C!0 N-times: 
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

for j in range(0, D!0):
  // sum!2[I], sum!3[I], sum!4[I] are vectors of size N  
  // Computes N intermediate sums "at once"
  sum!3[I] = PHI(sum!2[I], sum!4[I])       
  sum!4[I] = ADD_SIMD(sum!3[I],p!3[I,j])

@@

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion. 
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (essentially just ADD) has cost $\beta$ 
and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote 
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$. 
This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
in parallel still cost $2\alpha$. 
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

@@

\ana{Ishaq? Basically, define sequential schedule, then define an equivalent parallel schedule. A parallel schedule is equivalent if it preserves def-use relations in sequential schedule, or in other words, schedules def ahead of the use. Problem is to minimize cost of Parallel schedule.}

@@

\ana{Ishaq?}

@@

\begin{figure*}
  \includegraphics[width=1.0\linewidth]{figs_paper_SIMD/compiler_framework.png}
  \caption{Compiler Framework.} 
  \label{fig:compiler_framework}
\end{figure*}

%We assume arbitrarily nested loops in the MPC-source IR and read-only arrays. We assume that loops range from 0 to some constant N. Arrays are linearized (row-major order as in MOTION) and accesses are via functions of the induction variables of the enclosing loops.

\figref{fig:compiler_framework} presents an overview of our compiler. In this section, we describe several of the phases of the compiler. 
Sections~\secref{sec:vectorization} and~\secref{sec:div_and_conq} describe vectorization and divide-and-conquer.
We write $i,j,k$ to denote the loop nest: $i$ is the outermost loop, $j$, is immediately nested in $i$, and so on until $k$. 
We write $A[i,j,k]$ to denote canonical access to an array element. \ana{Need to state this more precisely.}

\subsection{Semantic Analysis}

\ana{Benjamin?}

@@

\subsection{Dependence Analysis}

\ana{TOTO Ana: Shorten this section. Lots of this is unnecessary.}

\subsubsection{Def-use Edges}

@@

\begin{itemize}
\item same-level forward $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest $i_1,i_2,...i_k$. E.g., $\texttt{d} = \mathit{SUB}(\texttt{S}[i,j], \texttt{C}[j])$ to $\texttt{p} = \mathit{MUL}(\texttt{d},\texttt{d})$ in Biometric is a same-level edge. A $\phi$ node can be a source of a same-level forward edge but not a target.
\item outer-to-inner forward $X\rightarrow Y$ where $X$ is in an outer loop nest, $i_1,i_2,...i_j$, and $Y$ is in an inner one, $i_1,i_2,...i_j,...,i_k$. A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer forward $X\rightarrow Y$ where $X$ is a $phi$-node in an inner loop nest, $i_1,i_2,...i_k,i_{k+1}$, and $Y$ is in the enclosing loop nest $i_1,i_2,...i_k$. E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer forward edge. Note that the source is \emph{always} a $\phi$-node in the immediately enclosing loop. The interpretation of this edge is that the use node $Y$ uses the definition made in the last iteration of the inner loop.
\ben{The current representation of pseudo $\phi$-nodes shows them attached to the loop header (i.e. in the loop).  We may want to clarify that they are also evaluated at the loop's termination.}
\item same-level back-edge $X\rightarrow Y$. $Y$ is a phi-node in the header of the loop and $X$ is a definition of the variable in the loop body. E.g., $\texttt{min}_1 = \mathit{MUX}(\texttt{c},\texttt{sum}_1,\texttt{min}_1)$ to $\texttt{min}_0 = \phi(\texttt{min}_1, 10000)$ in Biometric is a same-level back-edge.
\item inner-to-outer back-edge $X\rightarrow Y$: $X$ and $Y$ are both $\phi$-nodes for some variable. The source $X$ is in a loop nested into $Y$'s loop (not necessarily immediately).

@@

\subsubsection{Closures}

@@

\subsection{Taint Analysis}

\ana{Ben?}

\section{Vectorization}
\label{sec:vectorization}

\ana{TODO: Still work in progress, all over.}

%\ana{Add back-edges into Phase 1. A back-edge from a non-phi-node in loop $i$ to a phi-node in loop $i$'s header is a same-level edge. The only difference with the handling of normal same-level forward def-use is that the operand index will become $i-1$. A back edge from a phi-node in loop $j$ to a phi-node in loop $i$, where $j$ is nested in $i$ is an inner-to-outer edge and will require dropping dimension. We can show these are the only kinds of back-edges that may occur.}

\subsection{Raise Dimension and Drop Dimension}

@@

@@

%\ishaq{We need to think of a way to handle 4th case for def-use, when a nested block B def is followed by another, different, nested block B' use. Note: The handwritten notes say we can break it into 2 edges. One idea to this end is to insert a def' node which is a copy of the original def. The def' should be placed in a block that contains both B and B'.}
%\ishaq{THE ABOVE COMMENT IS RESOLVED. After transformation to SSA, there should be an intermediate variable (phi node) that is assigned a value depending on whether control branched or not. Should we put this into the writeup?}

@@

%\ana{Clarify why this helps. Still work in progress.}
This is to reduce the dimensionality of simd-ified computation. Technically, $\texttt{A}_j = \emph{MUX}(..., \texttt{A}_k, \texttt{A}_l)$ is
a simdified operation that can be carried out in parallel "in one round". However, particularly when \texttt{A} is a multi-dimensional array, there is
substantial increase in the size of the arrays (vectors) we send to SIMD operations. Refining to an update to a specific index would reduce
the size of those vectors. Note that this is a heuristic that handles a common case, but not all cases of array updates.
%\ana{TODO: Simplify this algorithm, taking into account the restriction to canonical updates. It should be handling \_all\_ cases.}

@@

\item same-level $X \rightarrow Y$. We do nothing, just propagate the array, which happens to be of the right dimension. 
%\ana{There might be some opportunities to do copy propagation optimization and save some cycles, but let's leave this for later.}

@@

\section{Divide-and-Conquer}
\label{sec:div_and_conq}
